Modeling the effect of aneuploidy on cancer evolution

### 2023-05-11 EvolutionaryRescue1.tex
Remus first update

The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.

@@

% TODO add subsection on the Gillespie simulations OK
\subsection*{Evolutionary simulation} 
Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.

At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
\end{flalign*}
\end{subequations}
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$. % TODO more details - you need to specify the events and how they change the number of cells. OK
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<1\%$. % TODO how do you determine the extinction prob? ref to equation OK
We continue running the simulation until:
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}
which is the population size at which the extinction probability of the population of mutant cell is less then $1\%$. 
% TODO do you use tau leap? if so, update the text accordingly. OK
When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}. If the increment is negative and larger then the subpopulation size then updated 
\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

@@

\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy3.pdf}
%\caption{}

@@

We write \cref{survprobw} as
\begin{equation}\label{survprobwapprox1}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
\begin{align}\label{aneuploidyresistentfirstapprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
\end{align}
\Cref{SurvPlot} shows that the survival probability, $p_{est}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases. % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario. OK
\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
and simulation results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
\label{SurvPlot}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
\end{align}

\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox2}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{AneuploidyDeleteriousApprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
\label{P_est}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}. }
\label{P_est}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid. The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}. For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue is due to initial aneuploid cell population.}
\label{FractionPlot}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

>>>>

The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are susceptible, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.

@@

% TODO add subsection on the Gillespie simulations OK
\subsection*{Evolutionary simulation} 
Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
\begin{figure}
\begin{center}
  \begin{tabular}{| l |c| c | c | c |}
    \hline
    Symbol & Name & Value & Units & References \\ \hline
    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
    $\lambda_w$ & Wildtype birth rate&  &   & 1 \\ \hline
    $\mu_w$ & Wildtype death rate&  &   & 1 \\ \hline
    $\lambda_a$ & Aneuploid birth rate&  &   & 1 \\ \hline
    $\mu_a$ & Aneuploid death rate&  &   & 1 \\ \hline
    $\lambda_m$ & Mutant birth rate&  &   & 1 \\ \hline
    $\mu_m$ & Mutant death rate&  &   & 1 \\ \hline
    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \cite{nowak2004evolutionary} \\ \hline
  \end{tabular}
\caption{.}
  \label{t1}
\end{center}
\end{figure}

At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
\end{flalign*}
\end{subequations}
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$. % TODO more details - you need to specify the events and how they change the number of cells. OK
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$. % TODO how do you determine the extinction prob? ref to equation OK
We continue running the simulation until:
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}
which is the population size at which the extinction probability of the population of mutant cell is less then $0.1\%$. 
% TODO do you use tau leap? if so, update the text accordingly. OK
When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}. If the increment is negative and larger then the subpopulation size then updated to be zero.
\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
%\caption{}

@@

We write \cref{survprobw} as
\begin{equation}\label{survprobwapprox1}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
\begin{align}\label{aneuploidyresistentfirstapprox}
p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
\end{align}
\Cref{SurvPlot} for $N=10^4$ and \Cref{SurvPlotLargeN} for $N=10^8$  show that the survival probability, $p_{resc}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases. % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario. OK
\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
and simulation results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.  Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
\label{SurvPlot}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.  Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
\label{SurvPlotLargeN}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!t]
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
%\label{SurvPlot}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}.
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{survprobresistantaneuploidy}
p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
\end{align}

\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
As a result, we can approximate:
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox2}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{AneuploidyDeleteriousApprox}
p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}
which we plot in Figure \ref{P_est} for initial population size $N=10^4$ and Figure \ref{P_est_large_N} for initial population size $N=10^8$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!t]
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
%\label{P_est}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}. }
\label{P_est}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{8}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_w-\mu_w$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations. }
\label{P_est_large_N}
\end{figure}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

In this paper, we have modelled a population of cancer cells which are exposed to chemotherapeutic drugs and decline towards extinction. Evolutionary rescue  is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. We have derived the probability of evolutionary rescue of the population of cancer cells under various demographic scenarious. The cancer cell population can escape extinction either through direct mutation or through mutation from aneuploidy. We have used multitype branching processes to study our model (Figure \ref{figureAneuploidy}) which allows us to obtain exact solutions for the probability of evolutionary rescue.

The case when the aneuploid cells are resistant can be approximated by the one step evolutionary rescue process where the aneuploidy rescues the population (Figure \ref{SurvPlot}). However, when the growth rate of the aneuploid cells is negative then they cannot rescue the population and they can only act as a stepping stone (Figure \ref{P_est}) through which the mutant can be obtain in a more expedient fashion, given that the aneuploid population declines slower then the wildtype population, compared to the case of direct mutation from the wildtype.

We observe from Figure \ref{RatioEvolRescue} that aneuploidy has a significant contribution towards evlutionary rescue. When aneuploidy is slightly increasing ($\Delta_a=10^{-3}$) the probability of evolutionary rescue is three orders of magnitude larger when aneuploidy is present compared to the case when aneuploidy is not present.

For our model we have assumed that cancer cell lineages are independent of each other. However this is not always true as cancer cells compete for resources which has an effect on the probability of evolutionary rescue (Figure \ref{SurvPlotNDataLogisticKComplete}). Future work should include using density dependent branching process in order to better model the conditions under cancer cells proliferate.

The presence of aneuploid cancer cells at the onset of chemotherapy can facillitate evolutionary rescue by acting as a stepping stone for the appearance of resistant mutant cells. From Figure \ref{FractionPlot} we observe that, for even a relative small fraction of the initial population being composed of aneuploid cells, evolutionary rescue is more likely to occour through the initial aneuploidy.

We propose experiments similar to the ones highlighted in \cite{martin2013probability} in order to test the predictions of our model. For example, in order to study the effects of initial population size on the probability of evolutionary rescue we propose to derive a large culture mass from a single cancer cell in permissive conditions and then dilute to a wide range of stating population sizes ($10^7-10^9$). Afterwards, we expose the population to anti-cancer drug which induces aneuploidy or to saline solution for control. Final density, in both cases, would be measured by optical density and the results compared to predictions from out model.

We observe from equations \eqref{survprobw} and \eqref{probresc} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
